---
title: "2024 WNBA Analysis"
author: "Michelle Hewson"
date: "2024-06-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Introduction
### This R file aims to analyze WNBA season data through various popular sports model frameworks, including the normal model, Bradley-Terry model, and Thurstone-Mosteller model. Team strengths will be estimated based on current season data and game outcome predictions for games that haven't been played yet will be calculated. These predictions will help inform sports analysts (and sports betters) for the upcoming WNBA playoffs.

## Read/process in data
```{r}
wnba <- read.csv("wnba_2024.csv")
```

```{r}
teams <- unique(wnba$HomeTeam)
num_teams <- length(teams)
```

## Normal model to estimate WNBA team strengths

```{r}
# Construct design matrix for normal model calculation
X <- outer(wnba$HomeTeam, teams, "==") - outer(wnba$AwayTeam, teams, "==")
W <- rbind(diag(num_teams-1), rep(-1, num_teams-1))

# Create schedule matrix
X_star <- X %*% W 
wnba$X_star <- X_star
ability_model <- lm(PtsDiff~X_star + AtHome + 0, data=wnba)
W_h <- rbind(cbind(W,0), c(rep(0,num_teams-1),1))

resid_stderr <- summary(ability_model)$sigma
team_ability_estimates <- W_h %*% ability_model$coefficients

V <- W_h %*% vcov(ability_model) %*% t(W_h)
dimnames(V) <- list(c(teams,"HFA"), c(teams,"HFA")) # Home field advantage

estimates <- data.frame(Estimate=team_ability_estimates, StdErr=sqrt(diag(V)))

row.names(estimates) <- c(teams,"Home Advantage")

# Ranking WNBA teams by normal strength estimates
team_ranks <- estimates[order(-estimates$Estimate),]
print(team_ranks)

# Write to csv for analysis in Python
write.csv(team_ranks, "normal_estimates_wnba2024.csv")
```
