---
title: "2024 WNBA Analysis"
author: "Michelle Hewson"
date: "2024-06-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Introduction
### This R file aims to analyze WNBA season data through various popular sports model frameworks, including the normal model, Bradley-Terry model, and Thurstone-Mosteller model. Team strengths will be estimated based on current season data and game outcome predictions for games that haven't been played yet will be calculated. These predictions will help inform sports analysts (and sports betters) for the upcoming WNBA playoffs.

## Read/process in data
```{r}
wnba <- read.csv("wnba_2024.csv")
```

```{r}
teams <- unique(wnba$HomeTeam)
num_teams <- length(teams)

# Construct design matrix for model calculations
X <- outer(wnba$HomeTeam, teams, "==") - outer(wnba$AwayTeam, teams, "==")
W <- rbind(diag(num_teams-1), rep(-1, num_teams-1))

# Create schedule matrix
X_star <- X %*% W 
wnba$X_star <- X_star
```

## Normal model to estimate WNBA team strengths

```{r}
# Normal model
normal_ability_model <- lm(PtsDiff~X_star + AtHome + 0, data=wnba)
W_h <- rbind(cbind(W,0), c(rep(0,num_teams-1),1))

resid_stderr <- summary(normal_ability_model)$sigma
team_ability_estimates <- W_h %*% normal_ability_model$coefficients

V <- W_h %*% vcov(normal_ability_model) %*% t(W_h)
dimnames(V) <- list(c(teams,"HFA"), c(teams,"HFA")) # Home field advantage

normal_estimates <- data.frame(Estimate=team_ability_estimates, StdErr=sqrt(diag(V)))

row.names(normal_estimates) <- c(teams,"Home Advantage")

# Ranking WNBA teams by normal strength estimates
team_ranks_normal <- normal_estimates[order(-normal_estimates$Estimate),]
print(team_ranks_normal)

# Write to csv for analysis in Python
write.csv(team_ranks, "normal_estimates_wnba2024.csv")
```
## Another approach: Thurstone-Mosteller model to estimate WNBA team strengths

```{r}
# Thurstone-Mosteller model (probit)
tm_model <- glm(HomeWin~X_star+AtHome+0,
family=binomial(link="probit"), data=wnba)

# Add home field advantage
W_h <- rbind(cbind(W,0),c(rep(0,num_teams-1),1)) 

# Team strength estimates
thurstone_estimates <- data.frame("Est"=W_h %*% tm_model$coefficients)
V.probit <- W_h %*% summary(tm_model)$cov.unscaled %*% t(W_h)
thurstone_estimates$StdErr <- sqrt(diag(V.probit))

# Format Thurstone-Mosteller dataframe
row.names(thurstone_estimates) <- c(teams,"HFA")
dimnames(V.probit) = list(c(teams,"HFA"), c(teams,"HFA"))
thurstone_estimates = thurstone_estimates[
order(-thurstone_estimates$Est),]
thurstone_estimates

# Write to csv for analysis in Python
write.csv(thurstone_estimates, "thurstone_estimates_wnba2024.csv")
```